Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market

### 2023-07-02 paper/reviewer_response.tex

I have elaborated to better explain this point in the relevant section of the Methods, as suggested by the reviewer.

Note also that in the \texttt{config.yaml} file for the pipeline, you can change the minimum mapping quality from 4 to 0 and re-run the pipeline.
If you do that, non-uniquely mapping reads are assigned randomly to one of the best-mapping reference species, which has the net effect (averaged over many reads) of giving a fractional score for these reads (eg, 0.5 for a read that maps to two species).
If you re-run the pipeline with this option, you will see that although the quantitative results shift some, the qualitative conclusions are largely unchanged.
}

4. I think the point of inconsistent calling of +/- SARS-CoV-2 samples by the original authors of the samples is VERY important and needs to be emphasized, especially because the positivity pattern was used to make very strong spatial correlation claims by Worobey and others. I would also suggest that instead of simply stating that 1/200M is statistically indistinguishable from 0/200M, it might be more informative to talk about limits of detection. In other words, if we accept 1/200M as a positive, we cannot rule out 0/200M as being positive, and perhaps talk about how many reads you would actually need to make some statistically sound claims (as a function of total read \#). You could put this in Table 1, for example.

\response{
I agree that the calling of samples as positive / negative based on the binary presence / absence of any SARS-CoV-2 read is not statistically well supported, as we cannot reject the null hypothesis that the underlying amount of SARS-CoV-2 is significantly lower in a sample with 0 in 200-million SARS-CoV-2 reads versus a sample with 1 in 200-million SARS-CoV-2 reads.
If we use a Fisher exact test and compare two samples with 200-million total reads one of which has 0 SARS-CoV-2 reads and the other of which has $N$ SARS-CoV-2 reads, we can only reject the null hypothesis that the two samples are drawn from reads with the same underlying distribution of SARS-CoV-2 content at $P \le 0.05$ when $N \ge 6$.

Therefore, as suggested by the reviewer, I have expanded on the relevant portion of the text explaining this fact, and described what threshold is needed for statistical significance on the statement that the SARS-CoV-2 content of the two samples is significantly different.

That said, I have not included any claims about how this fact would affect the conclusions that have been drawn by Worobey and others.
Certainly it is a factor that needs to be considered---however, many of the samples analyzed by Worobey and co-workers were also analyzed by RT-qPCR in addition to next-generation sequencing, so substantially more analysis of the full dataset and spatial distribution of samples is needed to determine how the above point might affect any conclusions by Worobey et al.
That analysis is beyond the scope of the current study, so here I have simply noted the lack of statistical rigor in how samples have been called positive / negative from deep-sequencing data by the Chinese CDC, but have refrained from speculating on how exactly that might impact downstream analyses using the Chinese CDC's positive / negative classifications.
}

### 2023-07-02 paper/paper.tex
Reviewer 2 point 4: consistency of calling samples + vs -

\begin{table}
\caption{
Inconsistency in criteria used to classify SARS-CoV-2 positivity in Chinese CDC study, illustrated with four example samples.
There is no consistent rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative: all three were reported negative by RT-qPCR, and A1 was not analyzed by sequencing while the difference in the number of SARS-CoV-2 reads between Q61 and E-10-29-2 is not statistically significant.
RT-qPCR results are from the second supplementary table of \citet{liu2023surveillance} (or equivalently the first table of \citet{liu2022surveillance}).
\label{tab:consistency}
}
{\footnotesize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llll}
\toprule
sample & \makecell[l]{RT-qPCR test \\ result (Ct)} & \makecell[l]{sequencing reads \\ mapping to SARS-CoV-2 \\ out of high-quality reads} & \makecell[l]{classification in \\ Chinese CDC study} \\
\midrule
F100 & positive (34.7) & 7,200 out of $2.6 \times 10^8$ & positive \\
Q61 & negative & 1 out of $2.1 \times 10^8$ & positive \\
E-10-29-2 & negative & 0 out of $1.9 \times 10^8$ & negative \\
A1 & negative & not sequenced & negative \\
\bottomrule
\end{tabular}
}
\end{table}

Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2.
However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
This inconsistency is illustrated in Table~\ref{tab:consistency} for four example samples: F100 is clearly positive (it both tested positive by RT-qPCR and contained thousands of reads mapping to SARS-CoV-2), but there is no statistical rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative.
See Table~\ref{tab:rt-qpcr} for merged sequencing and RT-qPCR data for all samples that were either sequenced or called as positive by the Chinese CDC.

>>>>

\begin{table}
\caption{
Inconsistency in criteria used to classify SARS-CoV-2 positivity in Chinese CDC study, illustrated with four example samples.
There is no consistent rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative: all three were reported negative by RT-qPCR, and A1 was not analyzed by sequencing while the difference in the number of SARS-CoV-2 reads between Q61 and E-10-29-2 is not statistically significant ($P = 1$ for both Fisher exact or $\chi^2$ test).
RT-qPCR results are from the second supplementary table of \citet{liu2023surveillance} (or equivalently the first table of \citet{liu2022surveillance}).
\label{tab:consistency}
}
{\footnotesize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llll}
\toprule
sample & \makecell[l]{RT-qPCR test \\ result (Ct)} & \makecell[l]{sequencing reads \\ mapping to SARS-CoV-2 \\ out of high-quality reads} & \makecell[l]{classification in \\ Chinese CDC study} \\
\midrule
F100 & positive (34.7) & 7,200 out of $2.6 \times 10^8$ & positive \\
Q61 & negative & 1 out of $2.1 \times 10^8$ & positive \\
E-10-29-2 & negative & 0 out of $1.9 \times 10^8$ & negative \\
A1 & negative & not sequenced & negative \\
\bottomrule
\end{tabular}
}
\end{table}

Now that the full data are available, we can examine the criteria used to classify samples as positive or negative for SARS-CoV-2.
It appears that the Chinese CDC classified samples as positive if they met \emph{either} of two criteria: they tested positive for SARS-CoV-2 by RT-qPCR, \emph{or} they were metagenomically sequenced and contained at least one read mapping to SARS-CoV-2 (Table~\ref{tab:rt-qpcr} gives sequencing and RT-qPCR data for all samples that were either sequenced or called positive by the Chinese CDC).
However, these criteria are not consistent because not all samples were analyzed by both methods, and the differences in the number of SARS-CoV-2 reads in the sequencing data are often not statistically significant between samples classified as positive versus negative.
This inconsistency is illustrated in Table~\ref{tab:consistency} for four example samples: F100 is clearly positive (it both tested positive by RT-qPCR and contained thousands of reads mapping to SARS-CoV-2), but there is no statistical rationale for classifying Q61 as positive but E-10-29-2 and A1 as negative.

To see the lack of statistical rationale for classifying Q61 as positive but E-10-29-2 as negative (Table~\ref{tab:consistency}), we can use a Fisher exact test (or alternatively a $\chi^2$ test) to test the null hypothesis that the reads in samples Q61 and E-10-29-2 are drawn from the same underlying distribution of viral and non-viral reads given that we observe 1 in $2.1 \times 10^8$ SARS-CoV-2 reads in Q61 and 0 in $1.9 \times 10^8$ SARS-CoV-2 reads in E-10-29-2.
Such an analysis shows that we cannot reject the null hypothesis that samples Q61 and E-10-29-2 have reads drawn from the same underlying distribution; this null hypothesis could only be rejected at a significance level of $P \le 0.05$ if sample Q61 contained $\ge$6 SARS-CoV-2 reads.
The lack of rationale is even more obvious when comparing Q61 and A1 (Table~\ref{tab:consistency}): both samples were negative by RT-qPCR, but Q61 was called positive based on metagenomic sequencing whereas A1 was never sequenced.
Therefore, if Q61 is called as positive then A1 should be called as indeterminate, since it was never subjected to the assay used to call Q61 as positive.
